Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-26 paper.tex
Tweaks to S3 benchmarks

\subsection{Extensibility: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for the Genomics England
data chromosome 2 described in Table~\ref{tab-genomics-england-data}.
an AWS S3 standard bucket. 
[TODO clarify what we're doing and remind reader]
We repeated the benchmarks above on the same subset
of 562,640 variants, using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask
required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s
occurs at 64 processes, with a corresponding
network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. Genotype data can therefore be fetched
from S3 around an order of magnitude more quickly than it can be decompressed
into RAM. [TODO QUESTION: can such rates be sustained, and be expected to 
happen in parallel? Do Durner et al consider this?]

@@

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5GiB/s using 128
processes. At this rate, the af-dist calculation for the target region
was completed in 4.15s (compare to 5.23min for the naive single-threaded
method above).
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: cost of c5.9xlarge per hour * 1.27
% ==                             $4.08 * 1.27
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.)

>>>>

\subsection{Extensibility: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for the Genomics England
data chromosome 2 described in Table~\ref{tab-genomics-england-data}
to an AWS S3 standard bucket. 
We repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s
occurs at 64 processes, with a corresponding
network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. 
% Richard Durbin asked whether we were just using up all of a scarce 
% resource on this point.
While these measurments are simplistic and may not be an 
accurate predictor of sustained bandwidth, it is still clear that 
genotype data can be fetched
from S3 around an order of magnitude more quickly 
than it can be decompressed into RAM. 
We have also not measured
the achievable bandwidth while scaling out across many instances,
but given the ubiquity of cloud-based services and the scale 
at which they operate, it seems reasonable to assume that the claimed
terabits per second of aggregate bandwidth from
S3~\cite{s3-performance-optimising} are achievable.

@@

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5GiB/s using 128
processes. At this rate, the af-dist calculation for the target region
was completed in 4.15s (compare to 5.23min for the naive single-threaded
method above).
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: cost of c5.9xlarge per hour * 1.27
% ==                             $4.08 * 1.27
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.
